---
title: "Analysis of the NYPD Shooting Incident Data"
date: "2021-07-31"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
This dataset lists every shooting incident in NYC between 2006 and the end of
last calendar year. As of the publishing date of this report, the data spans till
Dec 31, 2020. The dataset contains relevant information about the location, victims and the perpetrators as well. 

Load all the necessary libraries.
```{r load_library, message=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
theme_set(theme_minimal())
```

Download and load the NYC data.

```{r load_data}
df_url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

df <- read_csv(df_url, show_col_types = FALSE)

# Dimensions of the downloaded dataset
dim(df)
```

## Data Cleaning

```{r clean_data}
# Convert date from chr to Date format
df$OCCUR_DATE <- mdy(df$OCCUR_DATE) 

# Convert certain columns with categorical data from "chr" to factors
cols_to_factor <- c("BORO", "LOCATION_DESC", "PERP_SEX", "PERP_RACE", "VIC_SEX", "VIC_RACE")
df <- df %>% mutate_at(cols_to_factor, factor)

```

```{r drop_columns}
# Drop the columns that aren't necessary for the analysis
cols_to_drop <- c("X_COORD_CD", "Y_COORD_CD", "Latitude", "Longitude", "Lon_Lat")

clean_df <- df %>% select(-all_of(cols_to_drop))

# Let's check the number of missing values in each column
colSums(is.na(clean_df))
```
From the R output above we can see that columns related to perpetrator (PERP) are the ones that have the most amount of missing data. Perpetrators are the people who carry out a harmful/illegal act, in this case, they instigated the shooting incident based on the reports given by the victim/bystanders/crime reporters. I will not be dropping these columns

Additionally, the description of the place where the shooting incident took place is also missing from more than half of the dataset. We can exclude this column from our analysis.

For the 2 cases where the jurisdiction codes are misssing, we can safely drop those 2 rows since they represent a very (very) tiny portion of our dataset. 

```{r drop_na}
# Drop the location description column since its missing data for > 50% rows
clean_df <- clean_df %>% select(-LOCATION_DESC)

# Drop the 2 rows with missing jurisdiction codes
clean_df <- clean_df[!is.na(clean_df$JURISDICTION_CODE),]
```

## Data Analysis

```{r viz_1}
ggplot(data = clean_df, aes(x = year(OCCUR_DATE))) + geom_line(stat='count') + 
  expand_limits(y = 0) +  
  xlab("Year") + ylab("# of shooting incidents") + 
  ggtitle("# of shooting incidents in NYC per year") + 
  theme(plot.title = element_text(hjust = 0.5))
```
There was a steady decline in shooting incidents between 2011 and 2019 but the trend took a huge reversal in 2020. It reached some of the highest levels observed in the last 14 years. There could be a few reasons behind this:

1. Many businesses closed down amidst the pandemic. This might have lead to an increase in frustration among the general public. Higher frustration levels *might* have led to an increase in shooting incidents.

2. \# of gun sales might have increased as well. 

```{r viz_2, fig.width=12}
ggplot(data = clean_df, aes(x = fct_infreq(VIC_RACE))) + geom_bar() + 
  coord_flip() + 
  xlab("Victim's Race") + ylab("# of victims") + 
  ggtitle("# of shooting victims per race in NYC") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  geom_text(aes(label = ..count..), stat = "count", size = 4,
            fontface="bold", color="darkgreen", hjust=-0.01)
```

Black people are way more at risk of being victim to a shooting incident than all the other races combined.


```{r viz_3}
ggplot(data = clean_df, aes(x = fct_rev(fct_infreq(BORO)))) + geom_bar() +
  xlab("NYC Boroughs") + ylab("# of shooting incidents") +
  ggtitle("# of shooting victims per borough in NYC") + 
  theme(plot.title = element_text(hjust = 0.5))
```
Most of the shooting incidents took place in Brooklyn. But where did most of the murders take place? 

```{r viz_4}
clean_df[clean_df$STATISTICAL_MURDER_FLAG==TRUE, ] %>% 
  ggplot(aes(x = fct_rev(fct_infreq(BORO)))) + geom_bar() + 
  xlab("NYC Boroughs") + ylab("# of murders") +
  ggtitle("# of murders per borough in NYC") + 
  theme(plot.title = element_text(hjust = 0.5))
```
The distribution of murders is exactly similar to the previous visualization. 

## Data Modeling 
```{r model_df}
model_df <- clean_df %>% 
  group_by(year = year(OCCUR_DATE)) %>%
  summarise(total_incidents=n(), total_deaths=sum(STATISTICAL_MURDER_FLAG),
            .groups = 'drop')
head(model_df)
```
We've created a summary table of total incidents and total deaths in each year over the last 14 years. Lets create a simple linear model to see the correlation between # of incidents and # of deaths. 

```{r model}
mod <- lm(total_deaths ~ total_incidents, data=model_df)
summary(mod)
```
The R^2^ value is really high at 92% and the p-value is really small as well. This helps us conclude that the model can accurately predict the # of deaths based on the # of shooting incidents. 

```{r preds}
ggplot(data = model_df, aes(x = total_incidents, y = total_deaths)) +
  geom_point() + 
  stat_smooth(formula = y ~ x, method = "lm", col = "dodgerblue3", se=FALSE) +
  theme(panel.background = element_rect(fill = "white"),
        axis.line.x=element_line(), axis.line.y=element_line()) +
  xlab("# of incidents") + ylab("# of deaths") + 
  ggtitle("Comparing predictions vs actuals for # of deaths based on the linear model")
```
The blue line is made up of "predicted" deaths and is fairly linear since the model is *also* linear. The predictions are a little poor on both extremities (min and max). This could perhaps be further improved by looking at multiple linear regression: predicting # of deaths based on more than 1 variable 

## Conclusion

We've identified certain trends based on the data from the shooting incidents in NYC over the last 14 years:

- The borough with the highest numbers of shooting incidents *and* murders is Brooklyn. Further analysis could be performed here by correlating this information with population of the boroughs. 

- Black people are much more likely to be victims of shooting incidents than all the other races in the dataset combined. 

- \# of deaths and # of shooting incidents are highly correlated to each other. 

**Sources of bias**

- Poorer neighborhoods might have lower rate of reported incidents if the basic technological infrastrutuce isn't setup there. Poor cellphone signals, unscheduled power cuts, etc. 
- If people don't trust the police system, they are much less likely to report these incidents.

**Personal bias**

- As someone who dreams about living and working in Manhattan in the next few years, I was a little intrigued to dig deeper into Manhattan data exclusively and gather all the insights that I could.
- In order to mitigate this bias, I took a fair approach of analyzing **all** the boroughs at the same time to stay focused on the bigger picture.

